Mergeable Types

### 2017-02-25 paper/dali.tex
updates to intro
Programming Model.
dali.tex
Canvas concurrent sessions.
edits
Model edits.
opsem.
Sec. 4 beginning.

%% For double-blind review submission
\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Some recommended packages.
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{fullpage}
\usepackage{ifpdf}
\usepackage{graphicx}
%\usepackage[usenames,dvipsnames]{color}
\usepackage{subcaption}
\usepackage{stmaryrd}
%\usepackage[numbers]{natbib}
\usepackage{amsthm}
\usepackage{listings}          % format code
\usepackage{xspace}

@@

% Formatting commands
% -------------------
\newcommand{\code}[1]{\,{\tt #1}\,}
\newcommand{\spc}[0]{\quad}
\newcommand{\ALT}{~\mid~}
\newcommand{\rel}[1]{{R}_{\mathit{#1}}}
\newcommand{\conj}{~\wedge~}
\newcommand{\disj}{~\vee~}
\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
\newcommand{\ilrulelabel}[1]{{\sc #1}}
\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
                           {\begin{array}{c}#2\end{array}}}
\newcommand{\denot}[1]{\llbracket #1 \rrbracket}
\newcommand{\bind}{>\!\!>\!\!=}
\newcommand{\coloneqq}{::=}
\newcommand{\stepsto}{\longrightarrow}
\newcommand{\return}[1]{\C{return}\;#1}
\newcommand{\run}[2]{\C{run}\;#1\;#2}
\newcommand{\fork}[1]{\C{fork}\;#1}
\newcommand{\push}[1]{\C{push}\;#1}
\newcommand{\pull}{\C{pull}}
\newcommand{\semsucceq}{\succeq_{\circ}}
\newcommand{\sempreceq}{\preceq_{\circ}}
\newcommand{\under}[2]{#1\,\vdash\,#2}
\newcommand{\mbleto}{\rightsquigarrow_m}

% New colors
%------------
\definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}

% Listings Code
%---------------
\newcommand{\lstml}{
\lstset{ %
language=ML, % choose the language of the code
basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
keywordstyle=\color{Bittersweet},
% numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
% frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
commentstyle=\itshape\color{MidnightBlue},
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={module, @@deriving, not, : , /\\}
}}
\lstnewenvironment{ocaml}
    { % \centering
			\lstml
      \lstset{}%
      \csname lst@setfirstlabel\endcsname}
    { %\centering
      \csname lst@savefirstlabel\endcsname}

@@

%% Title information
\title[DaLi]{DaLi: Database as a Library}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

@@

\input{motivation}

@@

@@

%% Bibliography
%\bibliography{bibfile}

>>>>

%% For double-blind review submission
\documentclass[acmlarge,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Some recommended packages.
\usepackage{wrapfig}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{fullpage}
\usepackage{ifpdf}
\usepackage{graphicx}
%\usepackage[usenames,dvipsnames]{color}
\usepackage{subcaption}
\usepackage{stmaryrd}
%\usepackage[numbers]{natbib}
\usepackage{amsthm}
\usepackage{listings}          % format code
\usepackage{xspace}

@@

% Formatting commands
% -------------------
\newcommand{\code}[1]{\,{\tt #1}\,}
\newcommand{\spc}[0]{\quad}
\newcommand{\ALT}{~\mid~}
\newcommand{\rel}[1]{{R}_{\mathit{#1}}}
\newcommand{\conj}{~\wedge~}
\newcommand{\disj}{~\vee~}
\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
\newcommand{\ilrulelabel}[1]{{\sc #1}}
\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
                           {\begin{array}{c}#2\end{array}}}
\newcommand{\denot}[1]{\llbracket #1 \rrbracket}
\newcommand{\bind}{>\!\!>\!\!=}
\newcommand{\coloneqq}{::=}
\newcommand{\stepsto}{\longrightarrow}
\newcommand{\return}[1]{\C{return}\;#1}
\newcommand{\run}[2]{\C{run}\;#1\;#2}
\newcommand{\fork}[1]{\C{fork}\;#1}
\newcommand{\push}[1]{\C{push}\;#1}
\newcommand{\pull}{\C{pull}}
\newcommand{\semsucceq}{\succeq_{\circ}}
\newcommand{\sempreceq}{\preceq_{\circ}}
\newcommand{\under}[2]{#1\,\vdash\,#2}
\newcommand{\mbleto}{\rightsquigarrow_m}
\newcommand{\drawsome}{{\sf Canvas}\xspace}
\newcommand{\lang}{\lambda_b}
\newcommand{\length}[1]{\left|#1\right|}

% New colors
%------------
\definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}
\definecolor{MidnightBlue}{rgb}{0.0, 0.2, 0.4}

% Listings Code
%---------------
\newcommand{\lstml}{
\lstset{ %
language=ML, % choose the language of the code
basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
keywordstyle=\color{Bittersweet},
% numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
% frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
commentstyle=\itshape\color{MidnightBlue},
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={module, @@deriving, not, : , /\\}
}}
\lstnewenvironment{ocaml}
    { % \centering
			\lstml
      \lstset{}%
      \csname lst@setfirstlabel\endcsname}
    { %\centering
      \csname lst@savefirstlabel\endcsname}

@@

%% Title information
\title[]{Mergeable Types}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Distributed applications often replicate data to improve
  availability and fault tolerance.  But, the use of replication leads
  to a significantly more complex programming model.  To ensure a
  sensible measure of consistency among different replicas of an
  object, applications are typically structured so that the operations
  applied to a replicated object on one node are subsequently
  communicated to other nodes where they can be re-applied locally.
  As a result, the degree of restructuring and care necessary to
  transform a sequential or shared-memory concurrent application to a
  meaningful (replicated) distributed one can be substantial.

  In this paper, we present a new interpretation of replicated types
  that overcome these concerns.  Our declarative programming model
  called \name\ treats replication as a form of versioning, in which
  consistency is enforced through programmer-supplied \emph{merge}
  functions.  Distributed state is now viewed in terms of a tree of
  immutable object versions, with merge functions used implicitly by
  the underlying runtime system to periodically reconcile local
  versions to yield a consistent global state.  Thus, our model frees
  the programmer from having to think of low-level operational notions
  related to distribution and replication, and therefore does not
  entail any restructuring of application logic.

  We show how any ML datatype can be enriched with mergeable
  semantics, formalize a rich concurrency model based on such types ,
  and describe an instantiation of OCaml equipped with these features.
  Our implementation is integrated with a content-addressable
  distributed storage layer that allows any OCaml program to be easily
  transformed into an efficient, highly available, distributed
  application.
\end{abstract}

@@

\input{model}

@@

\input{mergeable-types}

@@

%% Bibliography
\bibliography{all}

### 2017-02-25 paper/intro.tex
Mergeable Type.
edits

\section{Introduction}

Real-world distributed programs are challenging to write and maintain
because they often conflate two distinct mechanisms.  The first
concerns the expression of application logic - how do we define
computations that are robust in the presence of distributed
communication among concurrently executing threads of control?  The
second deals with system concerns - how do we express notions of
visibility, replication, and consistency when the nodes participating
in such computations may be geographically distributed and the
networks that connect them unreliable?  To simplify reasoning in such
complicated environments, programming models often make strong
assumptions on the guarantees provided by an implementation (such as
serializability~\cite{Serializability} or strong
consistency~\cite{CDE+12,DNN+15}) that may inadvertently mask,
restrict, or ignore important albeit unpleasant realities (e.g.,
network partitions~\cite{Brewer2000,Gilbert2002}), or which may ignore
important system architectural features (e.g., replicated underlying
state~\cite{Walter}).  Given these unwarranted assumptions, program
behavior is often difficult to predict and verify.  On the other hand,
when these features are explicitly exposed to the programmer, e.g., by
requiring that applications be written in terms of specialized
distributed data structures~\cite{Burckhardt2014,,BFL+12,SPB+11} or
control primitives~\cite{Calm,Armstrong07}, simplicity, composability,
and ease-of-reasoning can suffer.

To illustrate how these kinds of problems manifest in practice,
consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic style in OCaml, with no special
reasoning principles needed to realize our desired functionality.  As
long as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior becomes significantly harder to understand.  For example, a
distributed implementation might wish to \emph{replicate} the counter
state on each node to improve latency (often referred to as
availability) or fault tolerance.  Unfortunately, replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose sigificant performance and availability penalties.  But,
without such heavyweight mechanisms, applying an \C{Add} operation on
one node may not be instanteously witnessed on another, which may in
turn be simultaneously attempting to perform its own \C{Add} or
\C{Mult} action.  While synchronizing the activities of all nodes to
ensure at most one such operation is performed at a time is
impractical, designating a single node to hold the counter state
eschewing replication altogether (as in a client-server
configuration), is also an undesirable solution, given its sensitivity
to network partitions and server failures.  Removig coordination
altogether by simply replicating counter state without having any
supporting consistency protocol is an equally infeasible approach.

A method often adopted~\cite{crdts, pldi15, gotsman-popl16} to address
this tension is to re-define a datatype's operations to return
\emph{effects} instead of values.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all replicas to
incorporate the effects of the original
operaton. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
with operations re-defined to returns effects.  Observe that the
\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
when applied at a replica (see \C{apply} in the figure), adds \C{x} to
the local counter value.  Note, however, that \C{add} and \C{mult} are
not commutative operations - assuming two replicas have the same
initial counter value of 0, applying the effect of adding 3 and then
multiplying 5 on one replica yields 15, while applying the effecting
of first multiplying 5 and then adding 3 yields 3 on the other.  Thus,
implementations must be carefully written to take the lack of
commutativity into account when defining how effects are applied at a
replica; in the figure, multiplication is expressed in terms of
addition to avoid the kind of undesirable behavior described above.

This implementation produces an eventually consistent global state in
which all replicas will eventually contain the same counter value.
While effects provide a low-level operational basis for handling
replicated state, the \emph{ad hoc} nature of the solution confounds
desirable high-level reasoning principles.  Indeed, the semantic gap
between the two versions of the counter, one cognizant of replication
and the other not, breaks backward compatibility.  More significantly,
its non-trivial construction must be developed in different guises for
every distributed data structure used by the application.
\begin{figure}
\begin{subfigure}[b]{0.4\textwidth}
  \begin{ocaml}
    module Counter: sig
      type t
      val add: int -> t -> t
      val mult: int -> t -> t
      val read: t -> int
    end = struct
      type t = int
      let add x v = v + (abs x)
      let mult x v = v * (abs x)
      let read v = v
    end
  \end{ocaml}
\caption{\C{Counter} library in OCaml}
\label{fig:counter-adt}
\end{subfigure}
\begin{subfigure}[b]{0.56\textwidth}
  \begin{ocaml}
    module Counter: sig
      type t
      type eff 
      val add: int -> t-> eff
      val mult: int -> t -> eff
      val apply: eff -> t -> t
      val read: t -> int
    end = struct
      type t = int
      type eff = Add of int
      let add x v = Add (abs x)
      let mult x v = Add (v * (abs x - 1))
      let apply (Add x) v = x + v
      let read v = v
    end
  \end{ocaml}
\caption{\C{Counter} library re-engineered for effect-based replication}
\label{fig:counter-rdt}
\end{subfigure}
\end{figure}
The lack of composability is another important downside of this
approach.  Consider an application that uses two replicated counters,
$c_1$ and $c_2$, bound by the invariant $c_2 \ge c_1$, duly enforced
by the application when updating $c_2$ or $c_1$.  An execution may
nonetheless witness anamolous states that violate the invariant
because updates to $c_1$ and $c_2$ may be applied independently in any
order on any replica.  For example, if a replica increments both
counters, a \C{read} operation performed at another replica may
witness the increment to $c_1$, but not $c_2$, thus violating the
invariant. The intention to atomically apply both effects cannot be
captured in the absence of external mechanisms to compose effects
(such as effect-based transactions~\cite{pldi15}).

On the other hand, a functional programmer is likely to approach this
problem very differently.  Rather than thinking about a low-level
operational variant of distributed state given in terms of explicit
effects as described above, a declarative interpretation of a
decentralized computation could be naturally described in terms of a
version tree whose root contains the initial counter value, and whose
branches represent different (immutable) versions maintained by
different nodes, with the state produced by a counter's operations on
a node recorded along the node's local branch.  Now, a globally
consistent view of a counter can be explained simply in terms of
\emph{merge} operations of different branch tips.  Treating
replication as merging yields a counter implementation that bears
strong similarity to the original sequential one:
  \begin{ocaml}
    module Counter = struct
      include Counter
      type mergeable_t = t [@@deriving persistence]
      let merge lca v1 v2 = lca + (v1 - lca) + (v2 - lca)
    end
  \end{ocaml}
The role of \C{lca} (least common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor state should not unwittingly duplicate the
contributions of the ancestor.  This interpretation of a replicated
datatype is thus given in terms of the evolution of a program state
impicitly associated with the different nodes that comprise a
distributed application with merge operations serving to communicate
and reconcile different local states.
\begin{figure}
  \begin{center}
    \includegraphics[scale=0.35]{Figures/dali-counter}
  \end{center}
  \caption{\small A multi-version counter can be expressed as a
    sequence of versions managed by logically-distributed concurrently
    executing threads.  Local versions produced by these threads
    periodically merge their state with one another.  In the figure,
    when the local version of Node 3, labeled 20, merges with the
    local version of Node 1, whose counter state is 15 at the time, a
    new counter value is produced.  This value takes into account the
    previous merged state (10) from which the two versions were
    derived to yield a new merged state of 25.  Observe that there is
    no explicit synchronization or coordination among versions.  In
    the figure, red circles represent state produced through merging,
    and blue circles represent local counter state produced by
    applying counter operations on a node-local version.  }
\end{figure}
The \name programming model realizes a monadic version control system
in OCaml centered around data, rather than file, coherence.  The \name
monad lets programmers write and compose concurrent computations
around multiple (implicit) versions of a mergeable datatype, an
ordinary datatype additionally equipped with a \emph{merge} function
responsible for deriving a consistent global state from a collection
of local versions of that state.  A computation makes progress by
forking concurrent versions of existing versions, creating later
versions along a \emph{branch}, or by merging branch tips.

While any OCaml datatype can be equipped with a merge operator, and
thus seamlessly participate in a distributed computation, instances of
datatypes whose merge operations are commutative with respect to their
local version state arguments (i.e., \C{v1} and \C{v2} in the counter
example above) are guaranteed to eventually converge to the same state
on all replicas.  For such types, OCaml programmers get the benefits
of realizing highly available (low-latency) distributed computation,
while continuing to enjoy the comforts of high-level reasoning and
the familiarity of using standard libraries.  Notably, while \name's
programming model makes no explicit reference to any specific
operational manifestation of a distributed system (e.g., programmers
do not need to worry about replicas and their management), we
demonstrate that it can be nonetheless efficiently realized on
existing geo-replicated distributed systems.

@@

    \item We describe the \name library that transparently adds
      persistence and replication features to a mergeable type. \name
      comes with a programming model that brings to bear the power and
      flexibility of a version control system to the administration of
      replicated data. \name hides the complexity of version control
      behind a monad, and exposes its functionality via a simple API
      that lets programmers define and compose concurrent ML
      computations around such data.  Issues related to replication
      only manifest implicitly in the definition of a merge operation
      that defines coherence among different instances of replicated
      state.

>>>>

\section{Introduction}
\label{sec:intro}

Real-world distributed programs are challenging to write and maintain
because they often conflate two distinct mechanisms.  The first
concerns the expression of application logic - how do we define
computations that are robust in the presence of distributed
communication among concurrently executing threads of control?  The
second deals with system concerns - how do we express notions of
visibility, replication, and consistency when the nodes participating
in such computations may be geographically distributed and the
networks that connect them unreliable?  To simplify reasoning in such
complicated environments, programming models often make strong
assumptions on the guarantees provided by an implementation (such as
serializability~\cite{Serializability} or strong
consistency~\cite{CDE+12,DNN+15}) that may inadvertently mask,
restrict, or ignore important albeit unpleasant realities (e.g.,
network partitions~\cite{Brewer2000,Gilbert2002}). Given these
unwarranted assumptions, program behavior is often difficult to
predict and verify.  On the other hand, when these features are
explicitly exposed to the programmer, e.g., by requiring that
applications be written in terms of specialized distributed data
structures~\cite{Burckhardt2014,BFL+12,SPB+11} or control
primitives~\cite{Calm}, simplicity, composability, and
ease-of-reasoning can suffer.

To illustrate how these kinds of problems manifest in practice,
consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic functional style, with no special
reasoning principles needed to realize desired functionality.  As long
as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instanteously witnessed on another, which may in the process of
simultaneously attempting to perform its own \C{Add} or \C{Mult}
action.  While synchronizing the activities of all nodes to ensure at
most one such operation is performed at a time is impractical,
designating a single node to hold the counter state eschewing
replication altogether (as in a client-server
configuration~\cite{Armstrong}), is also an undesirable solution,
given its sensitivity to network partitions and server failures, and
the negative performance impact it incurs in geo-distributed
environments~\cite{Walter}.  Removig coordination altogether by
simply replicating counter state without having any supporting
consistency protocol is an equally infeasible approach.

A method often adopted~\cite{SPB+11,Burckhardt2014} to address this
tension is to re-define a datatype's operations to return
\emph{effects} instead of values.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all replicas to
incorporate the effects of the original
operation. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
with operations re-defined to returns effects.  Observe that the
\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
when applied at a replica (see \C{apply} in the figure), adds \C{x} to
the local counter value.  Note, however, that \C{add} and \C{mult} are
not commutative operations - assuming two replicas have the same
initial counter value of 0, applying the effect of adding 3 and then
multiplying 5 on one replica yields 15, while applying the effect of
first multiplying 5 and then adding 3 yields 3 on the other.  Thus,
implementations must be carefully written to take the lack of
commutativity into account when defining how effects are applied at a
replica; in the figure, multiplication is expressed in terms of
addition to avoid the kind of undesirable behavior described above.

In this implementation, all replicas will eventually contain the same
counter value, assuming computation on the counter quiesces.  While
transmitting effects provide a low-level operational basis for
handling replicated state, the \emph{ad hoc} nature of the solution
confounds desirable high-level reasoning principles.  Indeed, the
semantic gap between the two versions of the counter, one cognizant of
replication and the other not, breaks backward compatibility.  Just as
significantly, its non-trivial construction must be developed in
different guises for every distributed data structure used by the
application.
\begin{figure}
\begin{subfigure}[b]{0.4\textwidth}
  \begin{ocaml}
    module Counter: sig
      type t
      val add: int -> t -> t
      val mult: int -> t -> t
      val read: t -> int
    end = struct
      type t = int
      let add x v = v + (abs x)
      let mult x v = v * (abs x)
      let read v = v
    end
  \end{ocaml}
\caption{\C{Counter} library in OCaml}
\label{fig:counter-adt}
\end{subfigure}
\begin{subfigure}[b]{0.56\textwidth}
  \begin{ocaml}
    module Counter: sig
      type t
      type eff 
      val add: int -> t-> eff
      val mult: int -> t -> eff
      val apply: eff -> t -> t
      val read: t -> int
    end = struct
      type t = int
      type eff = Add of int
      let add x v = Add (abs x)
      let mult x v = Add (v * (abs x - 1))
      let apply (Add x) v = x + v
      let read v = v
    end
  \end{ocaml}
\caption{\C{Counter} library re-engineered for effect-based replication}
\label{fig:counter-rdt}
\end{subfigure}
\end{figure}
The lack of composability is yet another important downside of this
approach.  Consider an application that uses two replicated counters,
$c_1$ and $c_2$, bound by the invariant $c_2 \ge c_1$, duly enforced
by the application when updating $c_2$ or $c_1$.  An execution may
nonetheless witness anamolous states that violate the invariant
because updates to $c_1$ and $c_2$ may be applied independently in any
order on any replica.  For example, if a replica increments both
counters, a \C{read} operation performed at another replica may
\begin{wrapfigure}{l}{.5\textwidth}
  \begin{center}
    \includegraphics[scale=0.4]{Figures/dali-counter}
  \end{center}
  \caption{\small A replicated counter can be expressed as a sequence
    of versions managed by logically-distributed concurrently
    executing threads.  Local versions produced by these threads
    periodically merge their state with one another.  In the figure,
    when the local version of Node 3, labeled 20, merges with the
    local version of Node 1, whose counter state is 15 at the time, a
    new counter value is produced.  This value takes into account the
    previous merged state (10) from which the two versions were
    derived to yield a new merged state of 25.  Observe that there is
    no explicit synchronization or coordination among versions.  In
    the figure, red circles represent counter state produced through
    merging, and blue circles represent state produced by applying
    counter operations on a node-local version.  }
\vspace*{-.85in}  
\end{wrapfigure}
witness the increment to $c_1$, but not $c_2$, thus violating the
invariant. The intention to atomically apply both effects cannot be
captured in the absence of external mechanisms to compose effects
(such as effect-based transactions~\cite{pldi15}).  Rather than
viewing a distributed computation in terms of \C{Add} effects produced
by operations, can we formulate a more declarative interpretation,
directly in terms of the counter value maintained by each node.  In
doing so, we observe that each replica essentially operates over its
own version of a counter.  Local operations on a replicated object can
be thought of as yielding new versions, thus producing a version tree,
with one branch for each replica.  Thus, each branch represents
different (immutable) versions maintained by different nodes, with the
state produced by a counter's operations on a node recorded along the
node's local branch.  Now, to generate a globally consistent view of a
counter, we only need to define a merge operation that explains how to
combine two local versions to produce a new version that reflects both
their states.

Framing replication as merging leads to a counter implementation that
bears strong similarity to the original sequential one:
  \begin{ocaml}
    module Replicated_Counter = struct
      include Counter
      type mergeable_t = t 
      let merge lca v1 v2 =
         lca + (v1 - lca) + (v2 - lca)
    end
  \end{ocaml}

The role of \C{lca} (least common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor state should not unwittingly duplicate the
contributions of the ancestor.  This interpretation of a replicated
datatype is thus given in terms of the evolution of a program state
impicitly associated with the different nodes that comprise a
distributed application with merge operations serving to communicate
and reconcile different local states.

In this paper, we develop a programming model that realizes a monadic
version control system centered around data, rather than file,
coherence.  The \name monad lets programmers write and compose
concurrent computations around multiple (implicit) versions of a
mergeable datatype, an ordinary ML datatype additionally equipped with
a \emph{merge} function responsible for deriving a consistent global
state from a collection of local versions of that state.  A
computation progresses by forking replicating new versions of existing
versions on geo-distributed nodes, creating new versions along a an
existing \emph{branch}, or by merging branch tips.

Our model allows ML programmers to get the benefits of realizing
highly available (low-latency) distributed computation, while
continuing to enjoy the comforts of high-level reasoning and the
familiarity of using standard libraries.  Notably, while \name's
programming model makes no explicit reference to any specific
operational manifestation of a distributed system (e.g., programmers
do not need to explicitly manage replicas), we demonstrate that it can
be nonetheless efficiently realized on existing geo-replicated
distributed systems.

@@

    \item We describe the \name library that transparently adds
      persistence and replication features to a mergeable type. \name
      comes with a programming model that brings to bear the power and
      flexibility of a version control system to the administration of
      replicated data. \name hides the complexity of version control
      behind a monad, and exposes its functionality via a simple API
      that lets programmers define and compose distributed ML
      computations around such data.  Issues related to replication
      only manifest implicitly in the definition of a merge operation
      that defines coherence among different instances of replicated
      state.
